A Horrocks-type theorem for even orthogonal $K_2$

### 2020-03-25 dl-paper.tex
Fixed some of the remaining issues with sections 1-4

Our main result is, thus, the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Let $A$ be a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all maps are injective:
\begin{equation} \nonumber
  \xymatrix{ \KO_2(2\ell, A) \ar[r] \ar[d] & \KO_2(2\ell, A[X]) \ar[d] \\ \KO_2(2\ell, A[X\inv]) \ar[r] & \KO_2(2\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion holds if one replaces the functor $\KO_2(2\ell, -)$ with $\K_2(\rD_\ell, -)$ or $\St(\rD_\ell, -)$. \end{theorem}
In the above statement $\KO_2(2\ell, -)$ denotes the unstable orthogonal $\K_2$-functor (see~\cref{sec:quillen}).

@@

\subsection{\texorpdfstring{$\K_2$}{K2}-groups and symbols} In our calculations we use two families of explicit elements of $\K_2(\Phi, R)$ called {\it Steinberg and Dennis--Stein symbols}. Notice that our notational conventions for symbols follow~\cite{DS73} and {\it not} more modern textbooks such as~\cite{Kbook}. In particular, the product of symbols is written multiplicatively.

@@

\begin{rem} It is possible to directly prove the stable symplectic counterpart of the above lemma. Indeed, we have the following chain of isomorphisms:
\[ \K_2(C_\infty, k[X, X^{-1}]) \cong \GW_2^2(k[X,X^{-1}]) \cong \GW_2^2(k) \oplus \GW_1^1(k) \cong \K_2^\mathrm{MW}(k) \oplus \K_1^\mathrm{MW}(k),\]
in which the first two are obtained from~\eqref{GW-concrete} and~\cref{bass-ft} and the last one follows from the isomorphism $\GW_1^1(k) \cong \K_1^\mathrm{MW}(k)$ of~\cite[Lemma~4.1.1]{AF17} and Matsumoto's theorem. 
Thus, we see that the result of Morita and Rehmann agrees with the modern literature on higher Grothendieck--Witt groups. \end{rem}

@@

Notice that $\KO_2(2n, R) = \Ker(\St(\rD_n, R) \to \mathrm{SO}(2n, R))$ contains $\K_2(\rD_n, R)$, but the converse is not generally true. On the other hand, from~\cref{lem:lowerKgroups} it follows that $\KO_3(2n, R)$ is isomorphic to the group $K_3^{G, P}$ in which $G = G(\rD_n, -) = \mathrm{Spin}(2n, -)$ and $P = \mathrm{Epin}(2n, -)$ is its elementary subfunctor.

@@

\begin{lemma} \label{lem:prop41}
Assume additionally that the residue field $k$ is of characteristic $\neq 2$.
Then the canonical map $f\colon C(\rD_\ell, B, I) \to C(\rD_\ell, R, I)$ is surjective for $\ell \geq 7$. \end{lemma}
\begin{proof}
Writing the starting portion of the homology long exact sequence for the map $\St(\rD_\ell, R) \to \St(\rD_\ell, R/I)$ 
and using the isomorphisms of~\cref{lem:relativeH3} and~\cref{lem:lowerKgroups} we obtain the following commutative diagram:
\begin{equation*}\xymatrix{
 \KO_3(2\ell, B) \ar[r] \ar[d] & \KO_3(2\ell, k[X]) \ar[d]_{f'} \ar@{->>}[r] & \ar[d]_{f} C(\rD_\ell, B, I) \\
 \KO_3(2\ell, R) \ar[r]        & \KO_3(2\ell, k[X, X^{-1}]) \ar@{->>}[r]        & C(\rD_\ell, R, I).}\end{equation*}
By~\cref{Panin-stability} theorem the map $f'$ can be identified with the canonical map $\GW_3^{[0]}(k[X]) \to \GW_3^{[0]}(k[X, X^{-1}])$.
By~\cref{bass-ft} $\GW_3^{[0]}(k[X, X^{-1}]) \cong \GW_3^{[0]}(k) \oplus \GW_2^{[3]}(k),$
but since the group $\GW_2^{[3]}(k)$ is trivial by~\cite[Lemma~2.2]{FRS12}, the map $f'$ (and hence $f$) is surjective.
\end{proof}

>>>>

Our main result is, thus, the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Let $A$ be a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all maps are injective:
\begin{equation} \nonumber
  \xymatrix{ \KO_2(2\ell, A) \ar[r] \ar[d] & \KO_2(2\ell, A[X]) \ar[d] \\ \KO_2(2\ell, A[X\inv]) \ar[r] & \KO_2(2\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion holds if one replaces the functor $\KO_2(2\ell, -)$ with $\K_2(\rD_\ell, -)$ or $\St(\rD_\ell, -)$. \end{theorem}
In the above statement $\KO_2(2\ell, -)$ denotes the unstable orthogonal $\K_2$-functor (see~\cref{sec:quillen}). Notice also that in the special case when $A=k$ is a field the above theorem follows from the results of U.~Rehmann, J.~Hurrelbrink and J.~Morita (see~\cref{field-injectivity} below).

@@

\subsection{\texorpdfstring{$\K_2$}{K2}-groups and symbols} In our calculations we use two families of explicit elements of $\K_2(\Phi, R)$ called {\it Steinberg and Dennis--Stein symbols}. Notice that our notational conventions for symbols follow~\cite{DS73} and {\it not} more modern textbooks such as~\cite{Kbook} (cf. with~Definition~III.5.11 ibid.). Notice also that the operation on $\K_2(\Phi, R)$ is always written multiplicatively.

@@

\begin{rem} It is possible to directly prove the stable symplectic counterpart of the above lemma (under the additional assumption that the characteristic of $k$ is not $2$). Indeed, we have the following chain of isomorphisms:
\[ \K_2(C_\infty, k[X, X^{-1}]) \cong \GW_2^2(k[X,X^{-1}]) \cong \GW_2^2(k) \oplus \GW_1^1(k) \cong \K_2^\mathrm{MW}(k) \oplus \K_1^\mathrm{MW}(k),\]
in which the first two isomorphisms are obtained from~\eqref{GW-concrete} and~\cref{bass-ft} and the last one follows from the isomorphism $\GW_1^1(k) \cong \K_1^\mathrm{MW}(k)$ of~\cite[Lemma~4.1.1]{AF17} and Matsumoto's theorem. 
Thus, we see that in the stable range the result of Morita and Rehmann agrees with the modern literature on higher Grothendieck--Witt groups. \end{rem}

@@

Notice that $\KO_2(2n, R) = \Ker(\St(\rD_n, R) \to \mathrm{SO}(2n, R))$ does not coincide with $\K_2(\rD_n, R) = \Ker(\St(\rD_n, R) \to \mathrm{Spin}(2n, R))$ but rather is related with it via the exact sequence
\[ \xymatrix{ 1 \ar[r] & \K_2(\rD_n, R) \ar[r] & \KO_2(2n, R) \ar[r] & \mu_2(R) \ar[r] & 1,} \]
which can be obtained from the exact sequence~\cite[p.~189]{Ba74} by means of the snake lemma (cf. also with~\cite[Corollary~4.3.5]{Ba74}).
On the other hand, from~\cref{lem:lowerKgroups} it follows that $\KO_3(2n, R)$ is isomorphic to the group $K_3^{G, P}$ in which $G = G(\rD_n, -) = \mathrm{Spin}(2n, -)$ and $P = \mathrm{Epin}(2n, -)$ is its elementary subfunctor.

@@

\begin{lemma} \label{lem:prop41}
Assume additionally that the residue field $k$ is of characteristic $\neq 2$.
Then the canonical map $f\colon C(\rD_\ell, B, I) \to C(\rD_\ell, R, I)$ is surjective for $\ell \geq 7$. \end{lemma}
\begin{proof}
Writing the starting portion of the homology long exact sequence for the map $\St(\rD_\ell, R) \to \St(\rD_\ell, R/I)$ 
and using the isomorphisms of~\cref{lem:relativeH3} and~\cref{lem:lowerKgroups} we obtain the following commutative diagram:
\begin{equation*}\xymatrix{
 \KO_3(2\ell, B) \ar[r] \ar[d] & \KO_3(2\ell, k[X]) \ar[d]_{f'} \ar@{->>}[r] & \ar[d]_{f} C(\rD_\ell, B, I) \\
 \KO_3(2\ell, R) \ar[r]        & \KO_3(2\ell, k[X, X^{-1}]) \ar@{->>}[r]        & C(\rD_\ell, R, I).}\end{equation*}
By~\cref{Panin-stability} the map $f'$ can be identified with the canonical map $\GW_3^{[0]}(k[X]) \to \GW_3^{[0]}(k[X, X^{-1}])$.
By~\cref{bass-ft} $\GW_3^{[0]}(k[X, X^{-1}]) \cong \GW_3^{[0]}(k) \oplus \GW_2^{[3]}(k),$
but since the group $\GW_2^{[3]}(k)$ is trivial by~\cite[Lemma~2.2]{FRS12}, the map $f'$ (and hence $f$) is surjective.
\end{proof}
